Concurrent Programming with CXL

### 2020-06-29 book/paper.tex
Latest
Better formatting

CXL is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
CXL is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming''~\cite{Schneider97}.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.  If not,
it reports the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the processes are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.

@@

\begin{problems}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:incpy} 1000 times.  How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \textit{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \textit{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{problems}

@@

Many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}%
\index{TAS}%
operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here $s$ points to a shared boolean variable and $p$
to a private boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
integers & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2], None} \\
\hline
Context & (generated by \textbf{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{e == e}, \texttt{e != e} & two CXL values of the same type \\
\hline
\texttt{e < e, e <= e} & any two CXL values \\
\hline
\texttt{e > e, e >= e} & any two CXL values \\
\hline
\texttt{e and e and ...} & two or more Booleans \\
\hline
\texttt{e or e or ...} & two or more Booleans \\
\hline
\texttt{not e} & a Boolean \\
\hline
\texttt{-e} & an integer \\
\hline
\texttt{e + e + ...} & two or more integers, sets, or lists \\
\hline
\texttt{e - e} & two integers or sets \\
\hline
\texttt{e * e * ...} & two or more integers or sets \\
\hline
\texttt{e / e} & two integers (integer division) \\
\hline
\texttt{e \% e} & two integers (division remainder) \\
\hline
\texttt{e~..~e} & two integers \\
\hline
\texttt{e [not] in e} & first is any CXL value, second is set \\
\hline
\texttt{e if e else e} & middle \texttt{e} must be a Boolean \\
\hline
\texttt{\^{}e} & an address \\
\hline
\texttt{choose e} & a set \\
\hline
\texttt{min e, max e} & a set \\
\hline
\texttt{cardinality e} & a set \\
\hline
\texttt{keys e} & a dictionary \\
\hline
\texttt{len e} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize e} & a bag \\
\hline
\texttt{atLabel e} & atom (corresponding to a label) \\
\hline
\texttt{nametag e} & \texttt{e} must be \texttt{()} \\
\hline
\texttt{processes e} & \texttt{e} must be \texttt{()} \\
\hline
\texttt{\&lv} & \texttt{lv} is an lvalue \\
\hline
\texttt{stop lv} & \texttt{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
\hline
\texttt{lv [op]= e} & op is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del lv} & delete\\
\hline
\texttt{assert b [, x]} & \texttt{b} is a Boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
\hline
\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
\hline
\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
\hline
\texttt{while b: S} & \texttt{b} is a Boolean, \texttt{S} a list of statements\\
\hline
\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
\hline
\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name tag & a dictionary with atoms \texttt{.name} and \texttt{.tag} \\
\hline
program counter & a Program Counter value pointing into the code \\
\hline
end & a Program Counter value indicating the end of the process code \\
\hline
atomic & if non-zero, the process is in atomic mode \\
\hline
stack & a list of CXL values \\
\hline
register & a CXL dictionary mapping atoms (representing process variables) to CXL values \\
\hline
stopped & a Boolean indicating if the context is stopped \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The name in a name tag is the name of the method that the process
is executing;
\item The tag in a name tag is its argument by default.  Optionally
\texttt{spawn} allows the tag to be specified explicitly;
\item A process terminates when it reaches the \texttt{Return} instruction
of the method or when it hits a fault.  Faults include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, as well as when an assertion fails;
\item The execution of a process in atomic mode does not get interleaved
with that of other processes.
\item The register of a process always contains a dictionary, mapping
atoms to arbitrary values.  The atoms correspond to the variable names
in a CXL program.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of CVM machine instructions \\
\hline
labels & a dictionary of atoms to program counters \\
\hline
variables & a dictionnary mapping atoms to values \\
\hline
ctxbag & a bag of non-stopped contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
failure & Boolean indicating some context has failed \\
\hline
initializing & Boolean indicating that the state is not fully initialized. \\
\hline
\end{tabular}
\vspace{1em}

>>>>

CXL is heavily influenced by Leslie Lamport's work on
TLA+, TLC, and PlusCal~\cite{Lamport02, Lamport09}.
CXL is designed to have a lower learning curve than those
systems, but is not as powerful.  When you finish this book
and want to learn more, we strongly encourage checking
those out.
Another excellent resource is Fred Schneider's book ``On
Concurrent Programming''~\cite{Schneider97}.
(This chapter is named after that book.)

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}%
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \textit{count} and
an array \textit{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \textit{self}.
It increments \textit{count} and sets \textit{done}[\textit{self}] to \texttt{True}.
Method \texttt{main} waits until \textit{done} flags are set to \texttt{True}.
After that, it verifies that the value of \textit{count} equals 2.
Like Python, CXL supports \texttt{assert} statements with two arguments:
the first is a Boolean condition, and the second is a value that is
reported if the condition fails, in this case
the actual value of \textit{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.
Note that although the processes are \emph{spawned} one at a time,
they will execute concurrently.  It is, for example, quite possible
that \texttt{incrementer(1)} finishes before \texttt{incrementer(0)}
even gets going.

@@

\begin{problems}
\item CXL programs can usually be easily translated into Python.  For example,
Figure~\ref{fig:incpy} is a Python version of Figure~\ref{fig:inc}.
\begin{enumerate}
\item Run Figure~\ref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run Figure~\ref{fig:incpy} 1000 times.
For example, if you are using the bash shell, you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
How many times does the assertion
fail (if any).
\end{enumerate}
\item Figure~\ref{fig:incmany} is a version of Figure~\ref{fig:incpy} that has each
incrementer thread increment \textit{count} \texttt{N} times.  Run Figure~\ref{fig:incpy}
10 times (using Python).
Report how many times the assertion fails and what the value of \textit{count}
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
\item Can you think of a fix to Figure~\ref{fig:inc}?  Try one or two different fixes
and run them through CXL.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
\end{problems}

@@

Many CPUs have an atomic ``test-and-set'' (TAS)
\index{test-and-set}%
\index{TAS}%
operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here $s$ points to a shared Boolean variable and $p$
to a private Boolean variable, belonging to some process.
The operation copies the value of the shared variable to the
private variable (the ``test'')
and then sets the shared variable to \texttt{True} (``set'').

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Boolean & \texttt{False}, \texttt{True} \\
\hline
Integer & \texttt{-inf, ..., -2, -1, 0, 1, 2, ..., inf} \\
\hline
Atom & \texttt{.example}, \texttt{.test1} \\
\hline
Program Counter & (method names are program counter constants) \\
\hline
Dictionary & \texttt{dict\{ .account: 12345, .valid: False \} } \\
\hline
Set & \texttt{\{ 1, 2, 3 \}, \{ False, .id, 3 \} } \\
\hline
Address & \texttt{\&lock, \&flags[2], None} \\
\hline
Context & (generated by \textbf{stop} expression) \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{e == e}, \texttt{e != e} & two CXL values of the same type \\
\hline
\texttt{e < e, e <= e} & any two CXL values \\
\hline
\texttt{e > e, e >= e} & any two CXL values \\
\hline
\texttt{e and e and ...} & two or more booleans \\
\hline
\texttt{e or e or ...} & two or more booleans \\
\hline
\texttt{not e} & a boolean \\
\hline
\texttt{-e} & an integer \\
\hline
\texttt{e + e + ...} & two or more integers, sets, or lists \\
\hline
\texttt{e - e} & two integers or sets \\
\hline
\texttt{e * e * ...} & two or more integers or sets \\
\hline
\texttt{e / e} & two integers (integer division) \\
\hline
\texttt{e \% e} & two integers (division remainder) \\
\hline
\texttt{e~..~e} & two integers \\
\hline
\texttt{e [not] in e} & first is any CXL value, second is set \\
\hline
\texttt{e if e else e} & middle \texttt{e} must be a boolean \\
\hline
\texttt{\^{}e} & an address \\
\hline
\texttt{choose e} & a set \\
\hline
\texttt{min e, max e} & a set \\
\hline
\texttt{cardinality e} & a set \\
\hline
\texttt{keys e} & a dictionary \\
\hline
\texttt{len e} & a dictionary (includes tuple or list) \\
\hline
\texttt{bagsize e} & a bag \\
\hline
\texttt{atLabel e} & atom (corresponding to a label) \\
\hline
\texttt{nametag e} & \texttt{e} must be \texttt{()} \\
\hline
\texttt{processes e} & \texttt{e} must be \texttt{()} \\
\hline
\texttt{\&lv} & \texttt{lv} is an lvalue \\
\hline
\texttt{stop lv} & \texttt{lv} is an lvalue \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{lv = e} & \texttt{lv} is an lvalue and \texttt{e} is an expression\\
\hline
\texttt{lv [op]= e} & op is one of \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/}, 
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del lv} & delete\\
\hline
\texttt{assert b [, x]} & \texttt{b} is a boolean.  If \texttt{False}, optionally print expression \texttt{e}\\
\hline
\texttt{const v = e} & \texttt{v} is an identifier, \texttt{e} is a constant expression\\
\hline
\texttt{def m(v, ...): S} & \texttt{m} and \texttt{v} are identifiers, \texttt{S} a list of statements\\
\hline
\texttt{let v = e, ...: S} & \texttt{e} is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{while b: S} & \texttt{b} is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{for v in e: S} & \texttt{e} is a set evaluated in order,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn p e [, t]} & \texttt{p} is a method, \texttt{e} is an expression, \texttt{t} is an optional tag (an expression) \\
\hline
\texttt{go c e} & \texttt{c} is a context, \texttt{e} is an expression \\
\hline
\texttt{import m} & \texttt{m} identifies a module \\
\hline
\end{tabular}
\vspace{1em}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
name tag & a dictionary with atoms \texttt{.name} and \texttt{.tag} \\
\hline
program counter & a Program Counter value pointing into the code \\
\hline
end & a Program Counter value indicating the end of the process code \\
\hline
atomic & if non-zero, the process is in atomic mode \\
\hline
stack & a list of CXL values \\
\hline
register & a CXL dictionary mapping atoms (representing process variables) to CXL values \\
\hline
stopped & a boolean indicating if the context is stopped \\
\hline
\end{tabular}
\vspace{1em}

Details:
\begin{itemize}
\item The name in a name tag is the name of the method that the process
is executing;
\item The tag in a name tag is its argument by default.  Optionally
\texttt{spawn} allows the tag to be specified explicitly;
\item A process terminates when it reaches the \texttt{Return} instruction
of the method or when it hits a fault.  Faults include divide by zero,
reading a non-existent key in a dictionary, accessing a non-existent
variable, as well as when an assertion fails;
\item The execution of a process in \emph{atomic mode} does not get interleaved
with that of other processes.
\item The register of a process always contains a dictionary, mapping
atoms to arbitrary values.  The atoms correspond to the variable names
in a CXL program.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
code & a list of CVM machine instructions \\
\hline
labels & a dictionary of atoms to program counters \\
\hline
variables & a dictionnary mapping atoms to values \\
\hline
ctxbag & a bag of non-stopped contexts \\
\hline
stopbag & a bag of stopped contexts \\
\hline
choosing & if not \texttt{None}, indicates a context that is choosing \\
\hline
failure & boolean indicating some context has failed \\
\hline
initializing & boolean indicating that the state is not fully initialized. \\
\hline
\end{tabular}
\vspace{1em}
